---
title: "GOV 51 Ps3"
author: "Jenny Jia"
date: "2026-01-04"
output: PDF_document
---
GitHub Repository: https://github.com/jennyjia319/gov51-ps3

# 1 Setup and Data Exploration
## 1.1 Load Packages and Data
```{r}
install.packages("glmnet")
library(tidyverse)
library(glmnet)
library(knitr)
compas <- read.csv("data/compas_clean.csv")
```
## 1.2 How many observations and variables are in the dataset? What is the unit of observation?
```{r}
dim(compas)
head(compas)
```
There are 7214 observations and 17 variables in this dataset. The unit of observation is defendants, since each row includes demographics and the criminal history of an individual. 

## 1.3 Examine the Outcome Variable
```{r}
mean(compas$recidivism)
```
The overall recidivism rate is 45.06515%, which means about 45% of the individuals were rearrested. The outcome is binary because it takes the value of either 0 or 1. This is an important consideration because predictions made from a linear model are interpreted as probabilities, so the model can produce values that are outside 0-1, which are invalid probabilities that are less fit for binary variables. 

## Summary Statistics
```{r}
compas |>
  summarise(across(everything(),
                   list(mean = mean, sd = sd),
                   na.rm = TRUE))
```
Variables like priors_count and age have the most variation, as shown by their high standard deviations. On the other hand, variables like native_american and asian show the least variation, with standard deviations close to zero. What this tells us is that their values are mostly concentrated around the binary value of 0 for the observations. 

## 1.5 Explore Key Predictors
```{r}
compas |>
  mutate(priors_groupped = case_when(
    priors_count == 0 ~ "0",        # no prior offenses
    priors_count <= 2 ~ "1–2",      
    priors_count <= 5 ~ "3–5",     
    priors_count <= 10 ~ "6–10",    
    TRUE ~ "11+"                   
  )) |>
mutate(priors_groupped = factor(priors_groupped, 
                          levels = c("0", "1–2", "3–5", "6–10", "11+"))) |> #  prior offenses years were out of order, fixed
  group_by(priors_groupped) |> # Group by the new categories |>
  
  summarise(rate = mean(recidivism)) |> # Calculate mean recidivism rate |>
  
  
  ggplot(aes(x = priors_groupped, y = rate)) +
  geom_col() +                      
  labs(x = "Prior Offenses",
       y = "Recidivism Rate",
       title = "Recidivism by Prior Offenses")
```
What we observe from the graph is that there is indeed a positive relationship between prior offenses and recidivism rates. This information suggests that prior offenses could be a good predictor of repeat offenses/recidivism. 

# OLS Baseline
## 2.1 Train/Test Split
```{r}
set.seed(51)
n <- nrow(compas)
train_idx <- sample(1:n, size = floor(0.8 * n))
train <- compas[train_idx, ]
test <- compas[-train_idx, ]

nrow(train)
nrow(test)
```
There are 5771 observations in the training set, and 1443 observations in the test set. 